---
title: "GelPrep Protein Covariate"
author: "BWE"
date: "Thursday, July 30, 2015"
output: 
  html_document:
    fontsize: 12pt
    css: "./CSSscripts/FloatTOC.css"
    toc: yes
---
```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Fig2/',
                      warning=FALSE, message=FALSE)
```

```{r}
require(limma)
```


# The Problem
GelPrep protein data does not have replication while Phosphopeptide data has nested replication. I would like to use the nested replicate via limma's duplicatecor function *while* fitting with GelPrep Prot as a covariate.


# The Data
```{r}
#create data matrix with replicated phospho data and single point estimate protein data
PhosProt3 <- merge(adata,GelPrep, by = "row.names")
row.names(PhosProt3) <- PhosProt3$Row.names
PhosProt3 <- PhosProt3[,2:16]
PhosProt3 <- as.matrix(PhosProt3)
dim(PhosProt3)
head(PhosProt3)
```

# The design matrix
```{r}
individual <- as.factor(c(rep("18486", times = 4), rep("18862", times = 4), rep("19160", times = 4), "18862", "18486", "19160"))  
Protein <- as.factor(c(rep(0,times = 12), rep(1, times = 3)))
SingleCaseGel <- data.frame(individual = individual, Protein = Protein)
row.names(SingleCaseGel) <- colnames(PhosProt3)
design_GelPrep <- model.matrix(~0 + individual + Protein, data = SingleCaseGel)
SingleCaseGel
design_GelPrep
```

# blocking and fitting
```{r}
block = c(1,1,2,2,3,3,4,4,5,5,6,6,0,0,0)
dupcor <- duplicateCorrelation(PhosProt3, design_GelPrep, block=block)
dupcor$consensus.correlation
fit <- lmFit(PhosProt3,design_GelPrep,block=block,correlation=dupcor$consensus)
contrast.matrix <- makeContrasts(individual18862-individual18486, individual19160-individual18862, 
                                 individual19160-individual18486, levels = design_GelPrep)
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)
```

#Open Questions
Overall I think this is the way it should be done but I am unsure about "0" blocks. Either the 'duplicateCorrelation' function ignores that data or it treats the 0s as a separate factor and includes them as an additional block. I think clearly the latter is happening given below:
```{r}
#note the design matrix for just the phospho data
design_base
#calculate dupcorr for the phosphodata alone
block2 = c(1,1,2,2,3,3,4,4,5,5,6,6)
dupcor2 <- duplicateCorrelation(adata,design_base,block=block2)
dupcor2$consensus.correlation == dupcor$consensus.correlation
```
#Options
Going forward we need a way to fit with GelPrep estimates while using technical replicate correlation information. This can either be done by altering the design matrix/block settings or by fitting using gls.series withing a for loop.



