---
title: "Variance components and protein copy numbers"
author: "Brett Engelmann"
date: "7/7/2015"
output: html_document
---
## Protein Copy Number Estimation - Zia's dataset

Copy number estimates from Zia's dataset are calculated within the *Perseus* sofware suite. This is an intensity based estimate (similar to iBAQ). Here the **raw** 'light' intensity channel is used for each of the three samples. For now no normalization is performed on intensity or output values.
```{r, echo=FALSE}
#input and explore copynumber data
Zianonorm <- read.table(file="./Perseus/TPcolumnind.txt", sep = "\t", header=T, fill = T, quote = "")#column independent (no normalization)
keepers <- grep("copy.", names(Zianonorm), value = T, ignore.case = T)
copynumbersall <- Zianonorm[,names(Zianonorm) %in% keepers]
copynumbers <- copynumbersall[,c(1,4,7)]
names(copynumbers) <- c("cn19160","cn18862","cn18486")
copynumbers <- log10(copynumbers)
boxplot(copynumbers, ylab = "log10(copynumber)")#looks better than cns1
```

The spearman correlation across samples is good. Scatterplots reveal (as expected) most of variation at low intensity end of distribution.
```{r, echo=FALSE, message=FALSE}
require(gplots)
require(Hmisc)
pairs(copynumbers)
cn.corr <- rcorr(as.matrix(copynumbers), type = "spearman")
heatmap.2(
  cn.corr$r,
  key = FALSE,
  dendrogram = "none",
  trace = "none",
  cellnote = round(cn.corr$r,2),
  notecex = 1.5,
  notecol = "black",
  col=bluered(25),
  cexCol=1.5,
  cexRow=1.5,
  margins = c(7,7)
)
```

Make an "S" curve of median(copynumberrank) vs median(log10(cn)) across the three samples.
```{r, echo=FALSE}
#get median copynumber and median rank, calculate relative rank.
cnMedian <- apply(as.matrix(copynumbers), 1, median)
Rank <- rank(1/cnMedian)
RelRank <- sapply(Rank,function(x) x/max(Rank))
#combine to create DF
MedCNDF <- data.frame(copynumber = cnMedian, Rank = Rank, RelRank = RelRank)
row.names(MedCNDF) <- Zianonorm$Majority.protein.IDs
par(mfrow = c(1,1))
plot(MedCNDF$RelRank, MedCNDF$copynumber)
```

## The continuum of protein concentrations is equally distributed across Variance Component clusters

Here I use the protein normalized/batch corrected variance component estimates in combination with the median copynumber estimates calculated from Zia's dataset as outlined above. Protein copynumber seems relatively evenly distributed across the variance component clusters.

```{r,echo=FALSE}
#protein prep sub to varcomp
proteinvarcompNames <- multExpanded1_withDE[multExpanded1_withDE$ppSubtoVarcomp == "+", "ppMajorityProteinIDs"]

##subset copynumber DF st only those subjected to NRE model are considered
MedCNDFvc <- MedCNDF[row.names(MedCNDF) %in% proteinvarcompNames, ]#1152

##VarcompProt formatting fix
VarcompProt$individual <- as.numeric(as.vector(VarcompProt$individual))
VarcompProt$biorep <- as.numeric(as.vector(VarcompProt$biorep))
VarcompProt$residual <- as.numeric(as.vector(VarcompProt$residual))

#add ppMajority IDs and copynumber information to VarcompProt
VarcompProt$ppMajorityIDs <- multExpanded1_withDE[multExpanded1_withDE$idmult %in% row.names(VarcompProt), "ppMajorityProteinIDs"]

##merge by medCNDFvc rownames
#rowNames <- row.names(VarcompProt)
VarcompProtMerged <- merge(VarcompProt,MedCNDFvc, by.x = "ppMajorityIDs", by.y = "row.names")
# row.names(VarcompProt) <- rowNames#lost rownames somehow. A few missing here.

#using red gray blue color gradient to represent copynumber relative rank
rgbpal <- colorRampPalette(c("Red","Gray","Blue"))

#plot final choice. RGB using relcnrank using 1 row and two columns
par(mfcol = c(1,2))
index <- length(unique(VarcompProtMerged$ppMajorityIDs))
plot(VarcompProtMerged$RelRank, VarcompProtMerged$copynumber, 
     col = rgbpal(index)[findInterval(VarcompProtMerged$RelRank,
                                      seq(range(VarcompProtMerged$RelRank)[1], 
                                          range(VarcompProtMerged$RelRank)[2], 
                                          length.out = index), rightmost.closed = T)], 
     pch = 19, ylab = "log10(copynumber)", xlab = "Relative Rank")

#quadrant plot with copynumber RelRank heatmap overlay
plot(log10(VarcompProtMerged$individual),log10(VarcompProtMerged$biorep),
     col = rgbpal(index)[findInterval(VarcompProtMerged$RelRank,
                                  seq(range(VarcompProtMerged$RelRank)[1], 
                                      range(VarcompProtMerged$RelRank)[2], 
                                      length.out = index), rightmost.closed = T)], 
     pch = 19, ylab = "log10 BioRep VarComp", xlab = "log10 Ind VarComp")
```


```{r}
sessionInfo()
```




